---
title: 'format ECHOR data'
author: 'Alex Huddell'
date: '10/7/2021'
output:
  word_document: default
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries and data

```{r load libs}
library(tidyverse)
library(lubridate)
library(here)


dl1<-read_csv(here('data','ECHOR_DL_1.csv'))
dl2<-read_csv(here('data','ECHOR_DL_1.csv'))


dat<-rbind(dl1,dl2)          

dat<-distinct(dat) #remove many duplicate rows

#read in CT data from N exchange program
CT<-read_csv(file=here("data","CT_NX_data.csv"))
CT<-distinct(CT) #remove a few duplicate rows
CT_permits<-unique(CT$permit)
```


#Data exploration visualization
```{r}
#analyzing distributions of parameter_desc codes
unique(dat$parameter_desc)

#remove unnecessary parameters
dat <- dat %>%
  filter(
    parameter_desc %in% c(
      "Flow rate",
      "Flow",
      "Flow, total",
      "Nitrogen, total [as N]",
      "Nitrogen, organic total [as N]",
      "Nitrogen, ammonia total [as N]",
      "Nitrogen, nitrite total [as N]",
      "Nitrogen, nitrate total [as N]" ,
      "Nitrogen, Kjeldahl, total [as N]",
      "Nitrogen, Total",
      "Ammonia [as N] + unionized ammonia",
      "Ammonia, unionized",
      "Nitrite + Nitrate total [as N]",
      "Ammonia nitrogen, total, [as N] 30 day",
      "Ammonia & ammonium- total",
      "Nitrogen, Ammonia  Total",
      "Nitrogen, Nitrite  Total",
      "Nitrogen, Nitrate  Total",
      "Nitrogen, Kjeldahl  Total",
      "Nitrogen, Kjeldahl, total [TKN] [water]",
      "Nitrogen, Total as NO3 + NH3",
      "Nitrogen, ammonia [NH3-N]",
      "Nitrogen, Total As N",
      "Nitrite Plus Nitrate Total",
      "Nitrogen, ammonia, total [as NH3]"
    )
  )

levels<-dat %>%
  group_by(parameter_desc) %>%
  tally()

ggplot(levels, aes(parameter_desc,n))+
  geom_col() +
  ylab("count of observations")+
  theme(axis.text.x=element_text(angle = 60, hjust=1))
  
#analyzing distributions of stats codes
levels2<-dat %>%
  group_by(statistical_base_short_desc) %>%
  tally()

ggplot(levels2, aes(statistical_base_short_desc,n))+
  geom_col() +
  ylab("count of observations")+
  theme(axis.text.x=element_text(angle = 60, hjust=1))

# #looking at distribution of designators
# levels(dat$esignator)

#already filtered to designator==1 when formatting the downloaded data

# levels<-dat %>%
#   group_by(Designator) %>%
#   tally()
# 
# ggplot(levels, aes(Designator,n))+
#   geom_col() +
#   ylab("count of observations")

#looking at distribution of monitoring locations
unique(dat$monitoring_location_desc)

levels<-dat %>%
  group_by(monitoring_location_desc) %>%
  tally()

ggplot(levels, aes(monitoring_location_desc,n))+
  geom_col() +
  ylab("count of observations")
```


#Standardize data

```{r pressure, echo=FALSE}

#remove leading zeroes from outfall
dat$outfall <- sub("^0+", "", dat$outfall)
dat$outfall

#reformating several columns
dim(dat)
names(dat)
unique(dat$parameter_desc)
param_codes<-as.data.frame(unique(dat$parameter_code))
unique(dat$monitoring_location_desc)
unique(dat$statistical_base_short_desc)
unique(dat$statistical_base_short_desc)

#renaming columns and 
dat<-dat%>%
  rename(value=dmr_value_standard_units,
         permit=npdes_id) %>%

# reformatting date and add a key and number of days in month column, then filter out some problematic units
  mutate(date=mdy(date)) %>%
  mutate(permit_outfall = paste0(permit, "_", outfall)) %>%
  mutate(key = paste0(permit, "_", outfall, "_", date)) %>%
  mutate(days_per_month = days_in_month(date)) %>%
   mutate(state=substr(permit,1,2)) %>% #add column with state abbreviation 
  mutate(permit_freq = paste0(permit, "_", `DMR Frequency of Analysis Code1`))

##########################
###NEED TO REVIEW ALL OF THESE AND ADAPT TO ECHOR DATA
#################################
  filter(`Monitoring Location` %in% c("1", "2", "EA", "EG", "Y")) %>% #this drops 5,945 keys and 95 permit/outfalls
  #removing infrequently used statistical base codes
  filter(
    `Statistical Base Long Desc` %in% c(
      "Monthly Avg.",
      "Average Value",
      "Daily Average",
      "30 Day Arithmetic Mean"
    )
  ) %>% #by this line, there are  57,703 keys dropped and 535 permit/outfalls dropped
  filter(!Parameter == "50047 - Flow, maximum during 24 hr period") %>% #by this line, there are 58,062 keys dropped and 538 permit/outfalls dropped
  filter(!permit %in% CT_permits) #%>% #this line removes the CT waterwater treatment plants because we will use CTDEEP data for those
  #by this line, there are 77,458 keys dropped and 613 permit/outfalls dropped
################################################

###AND WORK TO ADD ALL EXCEPTIONS
  #removing specific designators that cause duplicates--always removing the less frequent

#Then add
# drop_na(value) #drop rows with no data reported

############################
##################OLD CODE    
  # #removing unwanted discharge codes and monitoring locations
  # filter(monitoring_location_desc=="Effluent Gross") %>%
  # 
  # #removing infrequently used statistical base codes
  # filter(statistical_base_short_desc=="MO AVG" )%>%
  # 
  # #omitting this parameter_desc because we don't know what it means, but most
  # #observations report this AND 00056 flow rate,
  # #may want to include some of these observations later
  # filter(!param_codes %in% c("50050", "50047", "00059", "51061", "00625"))
  # need to go back and add this column back in?
  # filter(Designator=="1")

###################################
  
dat$parameter_desc <- droplevels(dat)$parameter_desc # dropping unused levels from dat$parameter_desc
dat$Units <- droplevels(dat)$Units # dropping unused levels from dat$parameter_desc
dat$statistical_base_short_desc <- droplevels(dat)$statistical_base_short_desc # dropping unused levels from dat$statistical_base_short_desc
dat$Designator <- droplevels(dat)$Designator # dropping unused levels from dat$Designator

dim(dat)
length(unique(as.character(dat$permit_outfall))) #XX permit/outfalls
length(levels(as.factor(dat$key))) #XX

```

In the next chunk, we reduce the number of units to three--one for flow, one for 
concentration, and another for loading rate. 

```{r}
#standardize units to kg per month, million L per month, or milligram per liter
dat <- dat %>% mutate(
  kg_per_month =
    case_when(
      Units == "(26) - Pounds per Day" ~ value / 2.205 * days_per_month,
      Units == "(01) - Kilograms per Day" ~ value * days_per_month, # converts per day to per month
      Units == "(35) - Grams per Day" ~ value / 1000 * days_per_month, # converts g to kg and per day to per month
      Units == "(2R) - Pounds per Hour" ~ value / 2.205 * 24 * days_per_month, # converts pounds to kg and hours to months
      Units == "(9T) - Million Pounds per Day" ~ value * 10^6 / 2.205 * days_per_month, # converts millions of pounds to kg per month
    )
)

#standardize units to millions of L per month
dat <- dat %>% mutate(
  million_L_per_month =
    case_when(
      Units == "(07) - Gallons per Day" ~ value / 10 ^ 6 * 3.785 * days_per_month,
      #converts from Gallons per day to millions L per month
      Units == "(08) - Cubic Feet per Second" ~ value * 28.317 / 10 ^ 6 * 60 * 60 * 24 * days_per_month,
      # converts cubic feet per second to million L per month
      Units == "(80) - Million Gallons per Month" ~ value * 3.785,
      # converts gallons to L
      Units == "(1X) - Gallons per Hour" ~ value / 10 ^ 6 * 3.785 * 24 * days_per_month,
      # converts gallons per hour to millions of L per month
      Units == "(8D) - Gallons per Month" ~ value / 10 ^ 6 * 3.785,
      # converts gallons to millions of L
      Units == "(03) - Million Gallons per Day" ~ value * 3.785 * days_per_month,
      # converts gallons to L and day to month
      Units == "(78) - Gallons per Minute" ~ value / 10 ^ 6 * 3.785 * 60 * 24 * days_per_month,
      # converts gallons to millions of L and minute to month
    )
)

#converting all N parameter_descs to mass terms of nitrogen and not others (i.e. ammonia, nitrite, etc.)


dat$value <-  ifelse(
  dat$parameter_desc == "00608 - Nitrogen, ammonia dissolved",
  dat$value * 14.0067 / 17.031,
  dat$value
)

dat$value <- ifelse(
  dat$parameter_desc == "34726 - Nitrogen, ammonia, total (as NH3)",
  dat$value * 14.0067 / 17.031,
  dat$value
)

#standardize concentrations to milligrams N per liter  
dat <- dat %>% mutate(
  mg_N_per_L=
    case_when(
   Units == "(19) - Milligrams per Liter" ~ value,
   Units == "(28) - Micrograms per Liter" ~ value / 1000, #convert from micrograms to milligram
    )
  )



```
Next, we reduce the number of nitrogen to the following: total nitrogen "TN",
organic nitrogen "ON", ammonia "NH3", nitrite "NO2", nitrate "NO3", and inorganic N "IN".
We reduce the flow parameter_descs to the following: flow rate 'flow_rate", 
maximum flow rate "flow_max_24h", total flow "flow_total", and dry flow "flow_dry".

```{r}
#cleaning up duplicate nitrogen and flow parameter_descs 
dat$param_simple <- recode(dat$parameter_desc,
    #Nitrogen parameter_descs
    "00600 - Nitrogen, total (as N)" = "TN",
    "00605 - Nitrogen, organic total (as N)" = "ON",
    "00610 - Nitrogen, ammonia total (as N)" = "NH3",
    "00615 - Nitrogen, nitrite total (as N)" = "NO2",
    "00620 - Nitrogen, nitrate total (as N)" = "NO3",
    "51445 - Nitrogen, Total" = "TN",
    "00608 - Nitrogen, ammonia dissolved" = "NH3",
    "00630 - Nitrite + Nitrate total (as N)" = "NO3NO2",
    "00609 - Ammonia nitrogen, total, (as N) 30 day" = "NH3",
    "49579 - Nitrogen, total Kjeldahl" = "TN",
    "51447 - Nitrogen, Nitrite  Total" = "NO2",
    "51448 - Nitrogen, Nitrate  Total" = "TN",
    "51449 - Nitrogen, Kjeldahl  Total" = "TN",
    "51446 - Nitrogen, Ammonia  Total" = "NH3",
    "51087 - Nitrogen, Kjeldahl, total (TKN) (water)" = "TN",
    "51489 - Nitrogen, Total as NO3 + NH3" = "NO3NH3",
    "34726 - Nitrogen, ammonia, total (as NH3)" = "NH3",
    "51085 - Nitrogen, ammonia (NH3-N)" = "NH3",
    "51425 - Nitrogen, Total As N" = "TN",
    "51450 - Nitrite Plus Nitrate Total" = "NO3NO2", 
    #flow parameter_descs
    "00056 - Flow rate" = "flow_rate",
    "74076 - Flow" = "flow_rate",
    "00058 - Flow rate" = "flow_rate",
    "82220 - Flow, total" = "flow_total",
    
  )

#analyzing distribution of parameter_descs
levels(dat$param_simple)

levels<-dat %>%
  group_by(param_simple) %>%
  tally()

ggplot(levels, aes(param_simple,n))+
  geom_col() +
  ylab("count of observations")


levels(dat$parameter_desc)

levels<-dat %>%
  group_by(parameter_desc) %>%
  tally()

ggplot(levels, aes(parameter_desc,n))+
  geom_col() +
  ylab("count of observations")+
  theme(axis.text.x=element_text(angle = 60, hjust=1))


```

#Standardizing statistics codes
For now,I excluded ALL but two stats because tackling them all together was overwhelming.
In the future, we may want to add back in some of the other statistics.

```{r}

dat$stat <- recode(dat$statistical_base_short_desc,
    "Monthly Avg." = "monthly_mean")

dat$stat <- droplevels(dat)$stat # dropping unused levels from dat$stat

levels(dat$stat)



```

This is an example of the duplicate values that  complicate reshaping the data from long to wide formatting. In this case, for key: CT0000086_1_1998-04-30.

```{r, echo=T}
#testing it out on test case to demonstrate issues with duplicates
test <- 
  dat %>%
  filter(key == "CT0000086_001_1998-04-30") 
test


#look at how many rows there are for each key
n_row<-dat %>%
  group_by(key)%>%
  tally() %>%
  arrange(n)

hist(n_row$n) #most observations have 1-2 rows only

```
Let's separate the keys for which kg N/month loads already exist from those which don't.

```{r}
#subset of data that report loads
N_load<- dat %>%
  group_by(key) %>%
  filter(!is.na(kg_per_month)) %>%
    #removing unwanted columns
  select(
    !c(value,
       parameter_desc,
      `Received Date`,
      monitoring_location_desc,
      Units,
      statistical_base_short_desc,
      Column,
      days_per_month,
      Designator,
      `Seasonal Indicator`,
      Limit,
      `DMR Frequency of Analysis Code1`,
      `Limit Sample Type`)
  ) %>%
  distinct() #removing any duplicate rows


#check if there are duplicates per key and one parameter_desc
dup<-N_load %>%
  group_by(key)%>%
  filter(param_simple=="TN") %>%
  tally() %>%
  arrange(desc(n))
dup #now there are no duplicates

#pivoting the TN and NH3 observations into wide format and removing flow and concentration columns
N_load_a<-N_load %>%
   pivot_wider(names_from = param_simple,
              values_from = kg_per_month) %>%
 select(-million_L_per_month,- mg_N_per_L,-stat ,-TKN,- NO3, - NO2, - NO3NO2, -state, -permit_freq, -permit_outfall_designator) %>%  rename(kg_N_TN_per_month=TN,
         kg_N_NH3_per_month=NH3)


names(N_load_a)

length(unique(as.character(N_load_a$permit_outfall))) #XX outfalls included here
```

Here we pivot the data wider calculate loads based on the concentrations and flow data, and remove the concentration and flow columns after we use them.


```{r}
#function to id columns with all NAs
not_all_na <- function(x) any(!is.na(x))


#review duplicates
keys_with_N<- dat %>%
  group_by(key) %>%
  filter(!any(!is.na(kg_per_month)))%>% #this removes any keys which do report in N loads
  filter(any(Parameter %in% c(
      "00600 - Nitrogen, total (as N)",
      "00605 - Nitrogen, organic total (as N)",
      "00610 - Nitrogen, ammonia total (as N)",
      "00615 - Nitrogen, nitrite total (as N)",
      "00620 - Nitrogen, nitrate total (as N)",
      "51445 - Nitrogen, Total",
      "00608 - Nitrogen, ammonia dissolved",
      "00630 - Nitrite + Nitrate total (as N)",
      "00609 - Ammonia nitrogen, total, (as N) 30 day",
      "49579 - Nitrogen, total Kjeldahl",
      "51447 - Nitrogen, Nitrite  Total",
      "51448 - Nitrogen, Nitrate  Total" ,
      "51449 - Nitrogen, Kjeldahl  Total" ,
      "51446 - Nitrogen, Ammonia  Total",
      "51087 - Nitrogen, Kjeldahl, total (TKN) (water)",
      "51489 - Nitrogen, Total as NO3 + NH3" ,
      "34726 - Nitrogen, ammonia, total (as NH3)",
      "51085 - Nitrogen, ammonia (NH3-N)",
      "51425 - Nitrogen, Total As N",
      "51450 - Nitrite Plus Nitrate Total"
    )) #this removes any keys which do report any other N parameters
  ) 
  
dup<-keys_with_N %>% 
  filter(param_simple == 'flow_rate') %>%
  group_by(key, param_simple)%>%
  drop_na('value')%>%
  tally() %>%
  filter(n>1)%>%
  arrange(desc(n))
dup #there are lots of flow rate duplicates from multiple parameters reported for the same date, but we will select flow parameters-- 50050 first, then 00056, then 74076, then 0058

dup<-keys_with_N %>%
  group_by(key) %>%
  filter(any(is.na(kg_per_month))) %>%
  filter(!param_simple == 'flow_rate') %>%
  group_by(key, param_simple)%>%
  drop_na('value')%>%
  tally() %>%
  arrange(desc(n))
dup #there are no more duplicates with N parameters



# here we separate different flow parameters to remove duplicates observations

#this filters the groups of rows at the grouped (key) level that do include the 50050  parameter
flow_50 <- keys_with_N %>%
   group_by(key) %>%
  filter(!any(!is.na(kg_per_month)))%>% #this removes any keys which do report in N loads
  # filter(is.na(kg_per_month)) %>%
  group_by(key) %>%
  filter(any(Parameter == '50050 - Flow, in conduit or thru treatment plant')) %>% #this filters a group of rows at the grouped (key) level, and selects first this most common parameter
  filter(Parameter == '50050 - Flow, in conduit or thru treatment plant')

#next, this filters the groups of rows at the grouped (key) level that do not include the 50050 parameter, and selects 00056
flow_56 <-keys_with_N %>%
  group_by(key) %>%
  filter(!any(!is.na(kg_per_month)))%>% #this removes any keys which do report in N loads
  # filter(is.na(kg_per_month)) %>%
  group_by(key) %>%
  filter(!any(Parameter == '50050 - Flow, in conduit or thru treatment plant')) %>%
  filter(Parameter == '00056 - Flow rate')

#next, this filters the groups of rows at the grouped (key) level that do not include the 50050 or 00056 parameter, and selects the second most common parameter 74076
flow_76 <- keys_with_N %>%
  group_by(key) %>%
  filter(!any(!is.na(kg_per_month)))%>% #this removes any keys which do report in N loads
  # filter(is.na(kg_per_month)) %>%
  group_by(key) %>%
  filter(!any(Parameter == '50050 - Flow, in conduit or thru treatment plant')) %>%
  filter(!any(Parameter == '00056 - Flow rate')) %>%
  filter(Parameter == '74076 - Flow')

#next, this filters the groups of rows at the grouped (key) level that do not include the 50050, 00056 or 74076 parameter, and selects the less common parameter 00058
flow_58 <- keys_with_N %>%
  group_by(key) %>%
  filter(!any(!is.na(kg_per_month)))%>% #this removes any keys which do report in N loads
  # filter(is.na(kg_per_month)) %>%
  group_by(key) %>%
  filter(!any(Parameter == '50050 - Flow, in conduit or thru treatment plant')) %>%
  filter(!any(Parameter == '00056 - Flow rate')) %>%
  filter(!any(Parameter == '74076 - Flow')) %>%
  filter(Parameter == '00058 - Flow rate')

#next, this filters the groups of rows at the grouped (key) level that do not include the 50050, 00056, 74076, or 00058 parameter, and selects the least common parameter 51061
flow_61 <- keys_with_N %>%
  group_by(key) %>%
  filter(!any(!is.na(kg_per_month)))%>% #this removes any keys which do report in N loads
  # filter(is.na(kg_per_month)) %>%
  group_by(key) %>%
  filter(!any(Parameter == '50050 - Flow, in conduit or thru treatment plant')) %>%
  filter(!any(Parameter == '00056 - Flow rate')) %>%
  filter(!any(Parameter == '74076 - Flow')) %>%
  filter(!any(Parameter == '00058 - Flow rate')) %>%
  filter(Parameter == '51061 - Flow (dry weather)')

#finally, we collect all nitrogen related parameters to bind to the flow observations
N_params<- keys_with_N %>%
group_by(key) %>%
  filter(!any(!is.na(kg_per_month)))%>% #this removes any keys which do report in N loads
    # filter(is.na(kg_per_month)) %>%
  filter(!Parameter == '50050 - Flow, in conduit or thru treatment plant') %>%
  filter(!Parameter == '00056 - Flow rate') %>%
  filter(!Parameter == '74076 - Flow') %>%
  filter(!Parameter == '00058 - Flow rate') %>%
  filter(!Parameter == '51061 - Flow (dry weather)') %>%
distinct() 

#bind all observations back together
flow_cleaned<-rbind(flow_50,flow_56,flow_76,flow_58,flow_61,N_params) 

#review duplicates
dups2<-flow_cleaned %>%
 # filter(is.na(kg_per_month)) %>%
  group_by(key, param_simple)%>%
  drop_na((value)) %>%
  tally() %>%
  arrange(desc(n))
dups2 #there are no more duplicates
```

Calculate loads for keys which do not report loads.

```{r}

#subset of data that DO NOT report loads, then
#pivot wider and then remove columns with only NAs (since some parameter_desc/unit combinations that don't make sense)
flow_N_concentration <- dat %>%
  group_by(key) %>%
  filter(any(is.na(kg_per_month))) %>%
   #removing unwanted columns
  select(
    !c(value,
       parameter_desc,
      `Received Date`,
      monitoring_location_desc,
      Units,
      statistical_base_short_desc,
      Column,
      days_per_month,
      Designator,
      `Seasonal Indicator`,
      Limit,
      `DMR Frequency of Analysis Code1`,
      `Limit Sample Type`)
  ) %>%
   pivot_wider(names_from = param_simple,
              values_from = c(mg_N_per_L, million_L_per_month)) %>%
  select_if(not_all_na) 

names(flow_N_concentration)

#calculating N loads
N_load_b <- flow_N_concentration %>%
  mutate(kg_N_TN_per_month = mg_N_per_L_TN * million_L_per_month_flow_rate) %>%
  mutate(kg_N_ON_per_month = mg_N_per_L_ON * million_L_per_month_flow_rate) %>%
  mutate(kg_N_NH3_per_month = mg_N_per_L_NH3 * million_L_per_month_flow_rate) %>%
  mutate(kg_N_NO2_per_month = mg_N_per_L_NO2 * million_L_per_month_flow_rate) %>%
  mutate(kg_N_NO3_per_month = mg_N_per_L_NO3 * million_L_per_month_flow_rate) %>%
   mutate(kg_N_TKN_per_month = mg_N_per_L_TKN * million_L_per_month_flow_rate) %>%
  mutate(kg_N_NO3NO2_per_month = mg_N_per_L_NO3NO2 * million_L_per_month_flow_rate) %>%
  mutate(kg_N_NO3NH3_per_month = mg_N_per_L_NO3NH3 * million_L_per_month_flow_rate) %>%
  select(
    -mg_N_per_L_TN,
    -mg_N_per_L_ON,
    -mg_N_per_L_NH3,
    -mg_N_per_L_NO2,
    -mg_N_per_L_NO3,
    -mg_N_per_L_NO3NO2,
    -mg_N_per_L_TKN,
    -mg_N_per_L_NO3NH3,
    -million_L_per_month_flow_rate,
    -stat
  )

#check if any rows contain both NO3 or NO2 and NO3NO2 before we combine them
filter(N_load_b, !is.na(kg_N_NO3_per_month) & !is.na(kg_N_NO3NO2_per_month))
filter(N_load_b, !is.na(kg_N_NO2_per_month) & !is.na(kg_N_NO3NO2_per_month))
#there are none, so we can add them without double counting

N_load_b$kg_N_sum_per_month <-
  ifelse(
    !is.na(N_load_b$kg_N_TKN_per_month),
    (
      N_load_b$kg_N_TKN_per_month + N_load_b$kg_N_NO3_per_month + N_load_b$kg_N_NO2_per_month + N_load_b$kg_N_NO3NO2_per_month
    ) ,
    (
      N_load_b$kg_N_NH3_per_month + N_load_b$kg_N_NO3_per_month + N_load_b$kg_N_NO2_per_month + N_load_b$kg_N_NO3NO2_per_month +        N_load_b$kg_N_ON_per_month
    )
  )


#remove intermediate columns we no longer need
N_load_b <- N_load_b %>%
  select(
    -kg_N_NH3_per_month,
    -kg_N_NO3_per_month,
    -kg_N_NO2_per_month,
    -kg_N_NO3NO2_per_month,
    -kg_N_ON_per_month,
    -kg_N_TKN_per_month
  )

#combine the different N columns; use total N when reported, or the sum of all other
#forms of N if TN is not reported
N_load_b$kg_N_TN_per_month <-
  ifelse(!is.na(N_load_b$kg_N_TN_per_month),
    N_load_b$kg_N_TN_per_month,
    N_load_b$kg_N_sum_per_month )
 
#remove other unnecessary columns
N_load_b<- N_load_b %>%
  select(-kg_N_sum_per_month, -permit_outfall_designator,- permit_freq, -state)

names(N_load_b)
length(unique(as.character(N_load_b$permit_outfall))) #138 outfalls included here

#for N_load_a
#combine the different N columns; use total N when reported, or the sum of all other
#forms of N if TN is not reported

N_load_a$kg_N_TN_per_month <-
  ifelse(
    !is.na(N_load_a$kg_N_TN_per_month),
    N_load_a$kg_N_TN_per_month,
    N_load_a$kg_N_NH3_per_month
  )

#remove NH3 column
N_load_a<- N_load_a %>%
  select(-kg_N_NH3_per_month)

names(N_load_a)

#combine dataframes
N_load<-rbind(N_load_a,N_load_b)


length(unique(as.character(N_load$permit_outfall))) #XX outfalls with/without locations
nrow(N_load) #XX

summary(N_load$kg_N_TN_per_month)

```

Format and add in the CT data

```{r}

CT<-read_csv(file=here("data","CT_NX_data.csv"))

CT<-distinct(CT) #remove a few duplicate rows

#formatting CT data to join with rest of ECHO data
CT_summary<-CT %>% 
  mutate(date=mdy(date)) %>%
  mutate(month_year=format(as.Date(date), "%Y-%m")) %>%
  mutate(days = day(date)) %>%
  mutate(Outfall=rep(1)) %>%
  mutate(kg_N_TN_per_period=N_load_lbs_day*days/2.205) %>%
  group_by(permit,Outfall,month_year) %>%
  summarise(kg_N_TN_per_month=sum(kg_N_TN_per_period)) %>%
  mutate(permit_outfall=paste0(permit,"_",Outfall)) %>%
  mutate(key = paste0(permit, "_", Outfall, "_", month_year)) 

names(CT)
permit_facility<-CT %>%
  group_by(permit,facility) %>%
  summarize(permit = first(permit),
            facility = first(facility))

CT_summary<-left_join(CT_summary, permit_facility)


N_load<- N_load %>%  
  mutate(month_year=format(as.Date(date), "%Y-%m")) %>%
  select(-date)

N_load$Outfall<-as.character(N_load$Outfall)
CT_summary$Outfall<-as.character(CT_summary$Outfall)



#join locations based on permit number and outfall number
N_load_CT_join <- rbind(N_load, CT_summary)

dim(N_load_CT_join) #43681     
dim(N_load) #25,745
dim(CT_summary) #17,936
length(unique(as.character(N_load_CT_join$permit_outfall))) #264 outfalls with locations
length(unique(as.character(N_load_CT_join$key))) #38,159 keys


```

Loading in NPDES database with NPDES ID to match locations and other data.

```{r}
outfall_locations <-
  read_csv(file = here("data", "NPDES_PERM_FEATURE_COORDS.csv"))

names(outfall_locations)

names(N_load_CT_join)

#need to remove leading zeroes from outfall column
N_load_CT_join$Outfall <- sub("^0+", "", N_load_CT_join$Outfall)
N_load_CT_join$Outfall

#create key variables with permit and locations for join
N_load_CT_join$permit_outfall <- paste0(N_load_CT_join$permit, "_", N_load_CT_join$Outfall)

outfall_locations$permit_outfall <-
  paste0(outfall_locations$EXTERNAL_PERMIT_NMBR,
         "_",
         outfall_locations$PERM_FEATURE_NMBR)


#join locations based on permit number and outfall number
dim(N_load_CT_join) #43,681, 7
length(unique(N_load_CT_join$permit_outfall)) #264 outfalls with/without locations
length(unique(N_load_CT_join$key)) #38,159 keys

N_load_out <- left_join(N_load_CT_join, outfall_locations, by = "permit_outfall")
dim(N_load_out) #44,626, 12

length(unique(as.character(N_load_out$permit_outfall))) #264 outfalls with locations
length(unique(as.character(N_load_out$key))) #38,159 keys

#there were rows added by the join (not sure why), but none subtracted

```

Now join the HUC8 IDs to the data.

```{r}
HUC8_outfall_locations <-
  read_csv(file = here('data', 'HUC8_NPDES_outfall_intersection.csv'))

#select columns
HUC8_outfall_locations <- HUC8_outfall_locations %>%
  select('EXTERNAL_PERMIT_NMBR',
         'PERM_FEATURE_NMBR',
         'huc8',
         'name')

names(HUC8_outfall_locations)

#need to remove leading zeroes from outfall column
HUC8_outfall_locations$PERM_FEATURE_NMBR <- sub("^0+", "", HUC8_outfall_locations$PERM_FEATURE_NMBR)
HUC8_outfall_locations$PERM_FEATURE_NMBR


#create key variables with permit and locations for join
HUC8_outfall_locations$permit_outfall <-
  paste0(HUC8_outfall_locations$EXTERNAL_PERMIT_NMBR,
         "_",
         HUC8_outfall_locations$PERM_FEATURE_NMBR)

#join locations based on permit number and outfall number
dat_join<- left_join(N_load_out, HUC8_outfall_locations, by = "permit_outfall")

dim(N_load_out) #53,599
dim(dat_join) #53,599
length(unique(as.character(dat_join$permit_outfall))) #264 outfalls with locations
length(unique(as.character(dat_join$key))) #38,159 keys

#the number of keys and outfalls with locations are the same as in the last step


write.csv(dat_join,
          file = here("data", "ECHO_data_clean.csv"),
          row.names = F)





```
