---
title: "format ECHO data"
author: "Alex Huddell"
date: "7/28/2021"
output:
  word_document: default
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries and data

```{r }
library(tidyverse)
library(lubridate)
library(here)

col_specifications<-('ccffffffffffddff') #compact string to specify column types

dat<-read_csv(file=here("data","Echo_LIS_complete_data.csv"), 
                     col_types = col_specifications)

dat<-distinct(dat) #remove a few duplicate rows

```
#Data exploration visualization
```{r}
#analyzing distributions of Parameter codes
levels(dat$Parameter)

levels<-dat %>%
  group_by(Parameter) %>%
  tally()

ggplot(levels, aes(Parameter,n))+
  geom_col() +
  ylab("count of observations")+
  theme(axis.text.x=element_text(angle = 60, hjust=1))
  
#analyzing distributions of stats codes
levels2<-dat %>%
  group_by(`Statistical Base Long Desc`) %>%
  tally()

ggplot(levels2, aes(`Statistical Base Long Desc`,n))+
  geom_col() +
  ylab("count of observations")+
  theme(axis.text.x=element_text(angle = 60, hjust=1))

#looking at distribution of designators
levels(dat$Designator)

levels<-dat %>%
  group_by(Designator) %>%
  tally()

ggplot(levels, aes(Designator,n))+
  geom_col() +
  ylab("count of observations")

#looking at distribution of monitoring locations
levels(dat$`Monitoring Location`)

levels<-dat %>%
  group_by(`Monitoring Location`) %>%
  tally()

ggplot(levels, aes(`Monitoring Location`,n))+
  geom_col() +
  ylab("count of observations")
```


#Standardize data

```{r pressure, echo=FALSE}
#impute zeros for  "No Discharge Code - C" 
dat$`Reported Value` <-
  ifelse(dat$`Reported Value` == "No Discharge Code - C" ,
         0,
         dat$`Reported Value`)

#reformating several columns
dim(dat)
names(dat)

#renaming columns and 
dat<-dat%>%
  rename(date=`Monitoring Period`,
        value=`Reported Value`,
         facility=`Facility Name`,
         permit=`Permit ID`) %>%

# reformatting date and add a key and number of days in month column, then filter out some problematic units
  mutate(date=mdy(date)) %>%
  mutate(key = paste0(permit, "_", Outfall, "_", date)) %>%
  mutate(days_per_month = days_in_month(date)) %>%
  
  #removing unwanted discharge codes and monitoring locations
  filter(!grepl("No Dishcharge Code",value))%>%
  filter(`Monitoring Location`=="1") %>%
  
  #removing infrequently used statistical base codes
  filter(`Statistical Base Long Desc`=="Monthly Avg." )%>%
  
  # omitting the few observations of these unclear units and/or Parameters
  filter(!Units == "(55) - Pounds") %>%
  filter(!Units == "(25) - Milliliters per Liter") %>%
  filter(!Parameter == "51450 - Nitrite Plus Nitrate Total") %>%

  #omitting this parameter because we don't know what it means, but most
  #observations report this AND 00056 flow rate,
  #may want to include some of these observations later
  filter(!Parameter=="50050 - Flow, in conduit or thru treatment plant") %>%
  filter(!Parameter=="50047 - Flow, maximum during 24 hr period") %>%
  filter(!Parameter=="00059 - Flow rate, instantaneous") %>%
  filter(!Parameter=="51061 - Flow (dry weather)") %>%
  
  #I'm not sure which designator is correct, but we'll limit analysis to 1 for now 
  #since it is most frequently reported
  filter(Designator=="1")

dat$Parameter <- droplevels(dat)$Parameter # dropping unused levels from dat$Parameter
dat$Units <- droplevels(dat)$Units # dropping unused levels from dat$Parameter
dat$Units <- droplevels(dat)$`Statistical Base Long Desc` # dropping unused levels from dat$`Statistical Base Long Desc`
dat$Designator <- droplevels(dat)$Designator # dropping unused levels from dat$Designator

dim(dat)

```

In the next chunk, we reduce the number of units to three--one for flow, one for 
concentration, and another for loading rate. 

```{r}
#standardize units to kg per month, million L per month, or milligram per liter
dat <- dat %>% mutate(
  kg_per_month =
    case_when(
      Units == "(26) - Pounds per Day" ~ value / 2.205 * days_per_month,
      Units == "(01) - Kilograms per Day" ~ value * days_per_month, # converts per day to per month
      Units == "(35) - Grams per Day" ~ value * 1000 * days_per_month, # converts g to kg and per day to per month
      Units == "(2R) - Pounds per Hour" ~ value / 2.205 * 24 * days_per_month, # converts pounds to kg and hours to months
      Units == "(9T) - Million Pounds per Day" ~ value * 10^6 / 2.205 * days_per_month, # converts millions of pounds to kg per month
    )
)

#standardize units to millions of L per month
dat <- dat %>% mutate(
  million_L_per_month =
    case_when(
      Units == "(07) - Gallons per Day" ~ value / 10 ^ 6 * 3.785 * days_per_month,
      #converts from Gallons per day to millions L per month
      Units == "(08) - Cubic Feet per Second" ~ value * 28.317 / 10 ^ 6 * 60 * 60 * 24 * days_per_month,
      # converts cubic feet per second to million L per month
      Units == "(80) - Million Gallons per Month" ~ value * 3.785,
      # converts gallons to L
      Units == "(1X) - Gallons per Hour" ~ value / 10 ^ 6 * 3.785 * 24 * days_per_month,
      # converts gallons per hour to millions of L per month
      Units == "(8D) - Gallons per Month" ~ value / 10 ^ 6 * 3.785,
      # converts gallons to millions of L
      Units == "(03) - Million Gallons per Day" ~ value * 3.785 * days_per_month,
      # converts gallons to L and day to month
      Units == "(78) - Gallons per Minute" ~ value / 10 ^ 6 * 3.785 * 60 * 24 * days_per_month,
      # converts gallons to millions of L and minute to month
    )
)

#converting all N parameters to mass terms of nitrogen and not others (i.e. ammonia, nitrite, etc.)


dat$value <-  ifelse(
  dat$Parameter == "00608 - Nitrogen, ammonia dissolved",
  dat$value * 14.0067 / 17.031,
  dat$value
)

dat$value <- ifelse(
  dat$Parameter == "34726 - Nitrogen, ammonia, total (as NH3)",
  dat$value * 14.0067 / 17.031,
  dat$value
)

#standardize concentrations to milligrams N per liter  
dat <- dat %>% mutate(
  mg_N_per_L=
    case_when(
   Units == "(19) - Milligrams per Liter" ~ value,
   Units == "(28) - Micrograms per Liter" ~ value / 1000, #convert from micrograms to milligram
    )
  )



```
Next, we reduce the number of nitrogen to the following: total nitrogen "TN",
organic nitrogen "ON", ammonia "NH3", nitrite "NO2", nitrate "NO3", and inorganic N "IN".
We reduce the flow parameters to the following: flow rate 'flow_rate", 
maximum flow rate "flow_max_24h", total flow "flow_total", and dry flow "flow_dry".

```{r}
#cleaning up duplicate nitrogen and flow parameters 
dat$param_simple <- recode(dat$Parameter,
    #Nitrogen parameters
    "00600 - Nitrogen, total (as N)" = "TN",
    "00605 - Nitrogen, organic total (as N)" = "ON",
    "00610 - Nitrogen, ammonia total (as N)" = "NH3",
    "00615 - Nitrogen, nitrite total (as N)" = "NO2",
    "00620 - Nitrogen, nitrate total (as N)" = "NO3",
    "00625 - Nitrogen, Kjeldahl, total (as N)" = "TN",
    "51445 - Nitrogen, Total" = "TN",
    "00608 - Nitrogen, ammonia dissolved" = "NH3",
    "00630 - Nitrite + Nitrate total (as N)" = "IN",
    "00609 - Ammonia nitrogen, total, (as N) 30 day" = "NH3",
    "49579 - Nitrogen, total Kjeldahl" = "TN",
    "51447 - Nitrogen, Nitrite  Total" = "NO2",
    "51448 - Nitrogen, Nitrate  Total" = "TN",
    "51449 - Nitrogen, Kjeldahl  Total" = "TN",
    "51446 - Nitrogen, Ammonia  Total" = "NH3",
    "51087 - Nitrogen, Kjeldahl, total (TKN) (water)" = "TN",
    "51489 - Nitrogen, Total as NO3 + NH3" = "IN",
    "34726 - Nitrogen, ammonia, total (as NH3)" = "NH3",
    "51085 - Nitrogen, ammonia (NH3-N)" = "NH3",
    "51425 - Nitrogen, Total As N" = "TN",
    "51450 - Nitrite Plus Nitrate Total" = "IN", 
    #flow parameters
    "00056 - Flow rate" = "flow_rate",
    "74076 - Flow" = "flow_rate",
    "00058 - Flow rate" = "flow_rate",
    "82220 - Flow, total" = "flow_total",
    
  )

#analyzing distribution of parameters
levels(dat$param_simple)

levels<-dat %>%
  group_by(param_simple) %>%
  tally()

ggplot(levels, aes(param_simple,n))+
  geom_col() +
  ylab("count of observations")


levels(dat$Parameter)

levels<-dat %>%
  group_by(Parameter) %>%
  tally()

ggplot(levels, aes(Parameter,n))+
  geom_col() +
  ylab("count of observations")+
  theme(axis.text.x=element_text(angle = 60, hjust=1))


```

#Standardizing statistics codes
For now,I excluded ALL but two stats because tackling them all together was overwhelming.
In the future, we may want to add back in some of the other statistics.

```{r}

dat$stat <- recode(dat$`Statistical Base Long Desc`,
    "Monthly Avg." = "monthly_mean")

dat$stat <- droplevels(dat)$stat # dropping unused levels from dat$stat

levels(dat$stat)



```

This is an example of the duplicate values that  complicate reshaping the data from long to wide formatting. In this case, for key: CT0000086_1_1998-04-30.

```{r, echo=T}
#testing it out on test case to demonstrate issues with duplicates
test <- 
  dat %>%
  filter(key == "CT0000086_001_1998-04-30") 
test


#look at how many rows there are for each key
n_row<-dat %>%
  group_by(key)%>%
  tally() %>%
  arrange(n)

hist(n_row$n) #most observations have 1-2 rows only

```
Let's separate the keys for which kg N/month loads already exist from those which don't.

```{r}
#subset of data that report loads
N_load<- dat %>%
  group_by(key) %>%
  filter(!is.na(kg_per_month)) %>%
    #removing unwanted columns
  select(
    !c(value,
       Parameter,
      `Received Date`,
      `Monitoring Location`,
      Units,
      `Statistical Base Long Desc`,
      Column,
      days_per_month,
      Designator,
      `Seasonal Indicator`,
      Limit,
      `DMR Frequency of Analysis Code1`,
      `Limit Sample Type`)
  ) %>%
  distinct() #removing any duplicate rows


#check if there are duplicates per key and one parameter
N_load %>%
  group_by(key)%>%
  filter(param_simple=="TN") %>%
  tally() %>%
  arrange(desc(n))
#now there are no duplicates

#pivoting the TN and NH3 observations into wide format and removing flow and concentration columns
N_load_a<-N_load %>%
   pivot_wider(names_from = param_simple,
              values_from = kg_per_month) %>%
  select(-million_L_per_month,- mg_N_per_L,-stat ) %>%
  rename(kg_N_TN_per_month=TN,
         kg_N_NH3_per_month=NH3)


```

Here we pivot the data wider calculate loads based on the concentrations and flow data, and remove the concentration and flow columns after we use them.


```{r}
#function to id columns with all NAs
not_all_na <- function(x) any(!is.na(x))

#subset of data that DO NOT report loads, then
#pivot wider and then remove columns with only NAs (since some parameter/unit combinations that don't make sense)
flow_N_concentration <- dat %>%
  group_by(key) %>%
  filter(is.na(kg_per_month)) %>%
   #removing unwanted columns
  select(
    !c(value,
       Parameter,
      `Received Date`,
      `Monitoring Location`,
      Units,
      `Statistical Base Long Desc`,
      Column,
      days_per_month,
      Designator,
      `Seasonal Indicator`,
      Limit,
      `DMR Frequency of Analysis Code1`,
      `Limit Sample Type`)
  ) %>%
   pivot_wider(names_from = param_simple,
              values_from = c(mg_N_per_L, million_L_per_month),
              values_fn = mean) %>%
  select_if(not_all_na) 

#calculating N loads
N_load_b <- flow_N_concentration %>%
  mutate(kg_N_TN_per_month = mg_N_per_L_TN * million_L_per_month_flow_rate) %>%
  mutate(kg_N_ON_per_month = mg_N_per_L_ON * million_L_per_month_flow_rate) %>%
  mutate(kg_N_NH3_per_month = mg_N_per_L_NH3 * million_L_per_month_flow_rate) %>%
  mutate(kg_N_NO2_per_month = mg_N_per_L_NO2 * million_L_per_month_flow_rate) %>%
  mutate(kg_N_NO3_per_month = mg_N_per_L_NO3 * million_L_per_month_flow_rate) %>%
  select(
    -mg_N_per_L_TN,
    -mg_N_per_L_ON,
    -mg_N_per_L_NH3,
    -mg_N_per_L_NO2,
    -mg_N_per_L_NO3,
    -million_L_per_month_flow_rate,
    -stat
  )

#combine N columns in N_load_b to standardize and join dataframes
names(N_load_b)


N_load_b$kg_N_inorganic_per_month <- N_load_b$kg_N_NH3_per_month +
  N_load_b$kg_N_NO3_per_month +
  N_load_b$kg_N_NO2_per_month

#adding organic N and inorganic N together as "composite" N
N_load_b$kg_N_sum_per_month <- N_load_b$kg_N_ON_per_month +
  N_load_b$kg_N_inorganic_per_month

#remove intermediate columns we no longer need
N_load_b <- N_load_b %>%
  select(
    -kg_N_NH3_per_month,-kg_N_NO3_per_month,
    -kg_N_NO2_per_month,
    -kg_N_inorganic_per_month,
    -kg_N_ON_per_month
  )

#combine the different N columns; use total N when reported, or the sum of all other
#forms of N if TN is not reported
N_load_b$kg_N_TN_per_month <-
  ifelse(
    !is.na(N_load_b$kg_N_TN_per_month),
    N_load_b$kg_N_TN_per_month,
    N_load_b$kg_N_sum_per_month
  )

#remove N sum column
N_load_b<- N_load_b %>%
  select(-kg_N_sum_per_month)

#for N_load_b
#combine the different N columns; use total N when reported, or the sum of all other
#forms of N if TN is not reported
N_load_b$kg_N_TN_per_month <-
  ifelse(
    !is.na(N_load_b$kg_N_TN_per_month),
    N_load_b$kg_N_TN_per_month,
    N_load_b$kg_N_sum_per_month
  )

names(N_load_b)

#for N_load_a
#combine the different N columns; use total N when reported, or the sum of all other
#forms of N if TN is not reported

N_load_a$kg_N_TN_per_month <-
  ifelse(
    !is.na(N_load_a$kg_N_TN_per_month),
    N_load_a$kg_N_TN_per_month,
    N_load_a$kg_N_NH3_per_month
  )

#remove NH3 column
N_load_a<- N_load_a %>%
  select(-kg_N_NH3_per_month)

names(N_load_a)

#combine dataframes
N_load<-rbind(N_load_a,N_load_b)

write.csv(N_load, file=here("data", "ECHO_data_clean.csv"),row.names = F)

```

