---
title: 'format PCS data'
author: 'Alex Huddell'
date: '10/7/2021'
output:
  word_document: default
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries and data

```{r load libs}
library(tidyverse)
library(lubridate)
library(here)

dat<-read_csv(file=here('data','PCS_LIS_watershed_data.csv'))


          
```


#Create key for each outfall/date and check how many rows per key.

```{r create key, echo=FALSE}
dat$date<-ymd(dat$`Monitoring Dt (MVDT)`)

dat <- dat %>%
  mutate(days_per_month = days_in_month(date)) %>%
  mutate(key = paste0(`Permit Num (NPID)`,`Discharge Num (DSCH)`,date))

nrow(dat) #there are 196,586 rows
dat %>% count(key) #there are 95,366 unique keys, so some keys have duplicate rows

#looking at one of they keys:
test<-dat %>% filter(key=='CT000008611989-01-31')

#plot the frequency of different parameters
dat$param<-as.factor(dat$'Parameter Cd (PRAM)')
levels(dat$param)

levels<-dat %>%
  group_by(param) %>%
  tally()

ggplot(levels, aes(param,n))+
  geom_col() +
  ylab('count of observations')+
  theme(axis.text.x=element_text(angle = 30, hjust=1))

```

```{r rename, organize, and filter}

#renaming columns and 
dat<-dat%>%
  rename(facility=`Facility Name Short (FNMS)`,
         permit=`Permit Num (NPID)`,
         Outfall=`Discharge Num (DSCH)`,
         quant_avg=`Meas Quant Avge (MQAV)`,
         conc_avg=`Meas Conc Avge (MCAV)`) %>%

# reformatting date and add a key and number of days in month column, then filter out some problematic units
  mutate(key = paste0(permit, '_', Outfall, '_', date)) %>%
  mutate(days_per_month = as.numeric(days_in_month(date))) %>%
  
  #removing unwanted discharge codes and monitoring locations
  filter(`Monitoring Location (MLOC)`=='1') %>%
  
    
  #removing one unclear unit that only relates to a few observations of flow maxima
  filter(!`Quant Unit Cd Desc (LQUCD)` =='(3R) - Million Gallons')%>%

#omitting this parameter because we don't know what it means, but most
  #observations report this AND 00056 flow rate,
  #may want to include some of these observations later
  filter(!param=='50050 - Flow, In Conduit Or Thru Treatment Plant') %>%
  filter(!param=='50047 - Flow, Maximum During 24 Hr Period') %>%
  filter(!param=='00059 - Flow Rate, Instantaneous') %>%
  filter(!param=='51061 - Flow (Dry Weather)') %>%
  filter(!param== '00625 - Nitrogen, Kjeldahl, Total (As N)') %>%
  #I'm not sure which designator is correct, but we'll limit analysis to 1 for now 
  #since it is most frequently reported
  filter(`Discharge Designator (DRID)`=='1')
  


dat$`Quant Unit Cd Desc (LQUCD)` <- droplevels(dat)$`Quant Unit Cd Desc (LQUCD)` # dropping unused levels from dat$`Quant Unit Cd Desc (LQUCD)`
dat$`Conc Unit Cd Desc (LCUCD)` <- droplevels(dat)$`Conc Unit Cd Desc (LCUCD)` # dropping unused levels from dat$`Quant Unit Cd Desc (LQUCD)`
dat$param<- droplevels(dat)$param # dropping unused levels from dat$param

#make values numeric
dat$quant_avg<-as.numeric(dat$quant_avg)
dat$conc_avg<-as.numeric(dat$conc_avg)


dim(dat)
```


In the next chunk, we reduce the number of units to three--one for flow, one for 
concentration, and another for loading rate. 

```{r standardize units}
#standardize units to kg per month, million L per month, or milligram per liter
dat <- dat %>% mutate(
  kg_per_month =
    case_when(
      `Quant Unit Cd Desc (LQUCD)` == '(26) - Pounds per Day' ~ quant_avg / 2.205 * days_per_month,
      `Quant Unit Cd Desc (LQUCD)` == '(01) - Kilograms per Day' ~ quant_avg  * days_per_month # converts per day to per month
             )
)

#standardize units to millions of L per month
dat <- dat %>% mutate(
  million_L_per_month =
    case_when(
      `Quant Unit Cd Desc (LQUCD)` == '(07) - Gallons per Day' ~ quant_avg / 10 ^ 6 * 3.785 * days_per_month,
      #converts from Gallons per day to millions L per month
      `Quant Unit Cd Desc (LQUCD)` == '(08) - Cubic Feet per Second' ~ quant_avg * 28.317 / 10 ^ 6 * 60 * 60 * 24 * days_per_month,
      # converts cubic feet per second to million L per month
      
      `Quant Unit Cd Desc (LQUCD)` == '(8D) - Gallons per Month' ~ quant_avg / 10 ^ 6 * 3.785,
      # converts gallons to millions of L
      `Quant Unit Cd Desc (LQUCD)` == '(03) - Million Gallons per Day' ~ quant_avg  * 3.785 * days_per_month,
      # converts gallons to L and day to month
      `Quant Unit Cd Desc (LQUCD)` == '(78) - Gallons per Minute' ~ quant_avg  / 10 ^ 6 * 3.785 * 60 * 24 * days_per_month,
      # converts gallons to millions of L and minute to month
    )
)

#converting all N parameters to mass terms of nitrogen and not others (i.e. ammonia, nitrite, etc.)


dat$conc_avg <-  ifelse(
  dat$param == '00608 - Nitrogen, ammonia dissolved',
  dat$conc_avg * 14.0067 / 17.031,
  dat$conc_avg
)

dat$conc_avg <- ifelse(
  dat$param == '34726 - Nitrogen, ammonia, total (as NH3)',
  dat$conc_avg * 14.0067 / 17.031,
  dat$conc_avg
)

dat$conc_avg <- ifelse(
  dat$param ==  '71850 - Nitrogen, Nitrate Total (As No3)' ,
  dat$conc_avg * 14.0067 / 62.0049,
  dat$conc_avg
)

   
#standardize concentrations to milligrams N per liter  
dat <- dat %>% mutate(
  mg_N_per_L=
    case_when(
   `Conc Unit Cd Desc (LCUCD)` == 'Milligrams Per Liter' ~ conc_avg,
   `Conc Unit Cd Desc (LCUCD)` == 'Micrograms Per Liter' ~ conc_avg / 1000, #convert from micrograms to milligram
    )
  )



```
Next, we reduce the number of nitrogen to the following: total nitrogen 'TN',
organic nitrogen 'ON', ammonia 'NH3', nitrite 'NO2', nitrate 'NO3', and inorganic N 'IN'.
We reduce the flow parameters to the following: flow rate 'flow_rate', 
maximum flow rate 'flow_max_24h', total flow 'flow_total', and dry flow 'flow_dry'.

```{r simplify parameters}
#cleaning up duplicate nitrogen and flow parameters 
dat$param<-as.factor(dat$param)

         
#dat<-dat %>% mutate(param_simple=recode(dat$param,
#cleaning up duplicate nitrogen and flow parameters 
dat$param_simple <- recode(dat$param,

    # #Nitrogen parameters
    '00600 - Nitrogen, Total (As N)' = 'TN',
    '00605 - Nitrogen, Organic Total (As N)' = 'ON',
    '00608 - Nitrogen, Ammonia Dissolved' = 'NH3',
    '00610 - Nitrogen, Ammonia Total (As N)' = 'NH3',
    '00615 - Nitrogen, Nitrite Total (As N)' = 'NO2',
    '00620 - Nitrogen, Nitrate Total (As N)' = 'NO3',
    '00630 - Nitrite Plus Nitrate Total 1 Det. (As N)' = 'IN',
    '34726 - Nitrogen, Ammonia, Total (As Nh3)' = 'NH3',
    '49579 - Nitrogen, Total Kjeldahl' = 'TN',
    '51087 - Nitrogen, Kjeldahl, Total (Tkn) (Water)' = 'TN',
    '71850 - Nitrogen, Nitrate Total (As No3)' ='NO3',
    #flow parameters
    '00056 - Flow Rate' = 'flow_rate',
    '74076 - Flow' = 'flow_rate',
    '00058 - Flow Rate' = 'flow_rate',
    '82220 - Flow, Total' = 'flow_total',
    
  )

#analyzing distribution of parameters
levels(dat$param_simple)

levels<-dat %>%
  group_by(param_simple) %>%
  tally()

ggplot(levels, aes(param_simple,n))+
  geom_col() +
  ylab('count of observations')


levels(dat$param)

levels<-dat %>%
  group_by(param) %>%
  tally()

ggplot(levels, aes(param,n))+
  geom_col() +
  ylab('count of observations')+
  theme(axis.text.x=element_text(angle = 60, hjust=1))


```

Let's separate the keys for which kg N/month loads already exist from those which don't.

```{r separate observations reporting loads}
#review duplicates
dat %>%
  filter(!is.na(kg_per_month)) %>%
  filter(param_simple=='TN') %>%
  group_by(key)%>%
  tally() %>%
  arrange(desc(n))

#subset of data that report loads
N_load<- dat %>%
  group_by(key) %>%
  filter(!is.na(kg_per_month)) %>%
    #removing unwanted columns
  select(
    !c('Monitoring Dt (MVDT)',
        'Discharge Designator (DRID)', 
       'Parameter Cd (PRAM)',
       'Monitoring Location (MLOC)',
       '#MULTIVALUE',   
       'Meas Quant Maxm (MQMX)', 
       'Quant Unit Cd (LQUC)',
       'Quant Unit Cd Desc (LQUCD)',
       'Meas Conc Minm (MCMN)',
       'Meas Conc Maxm (MCMX)...15',
       'Meas Conc Maxm (MCMX)...16',
       'Conc Unit Cd (LCUC)',
       'Conc Unit Cd Desc (LCUCD)' )
  ) %>%
  distinct() #removing any duplicate rows


#check if there are duplicates per key and one parameter
N_load %>%
  group_by(key)%>%
  filter(param_simple=='TN') %>%
  tally() %>%
  arrange(desc(n))
# there are no duplicates

#pivoting the TN and NH3 observations into wide format and removing flow and concentration columns
N_load_a<-N_load %>%
   pivot_wider(names_from = param_simple,
              values_from = kg_per_month) %>%
  select(-million_L_per_month,- mg_N_per_L ) %>%
  rename(kg_N_TN_per_month=TN,
         kg_N_NH3_per_month=NH3)


```

Here we pivot the data wider calculate loads based on the concentrations and flow data, and remove the concentration and flow columns after we use them.


```{r calculating loads for other observations}
#function to id columns with all NAs
not_all_na <- function(x) any(!is.na(x))

#review duplicates
dat %>%
  filter(is.na(kg_per_month)) %>%
  group_by(key, param_simple)%>%
  tally() %>%
  arrange(desc(n))

#many parameters have duplicates that we need to resolve before pivoting

#separating different flow parameters to remove duplicates

flow_56 <- dat %>%
  filter(is.na(kg_per_month)) %>%
  group_by(key) %>%
  filter(any(param == '00056 - Flow Rate')) %>%
  filter(param == '00056 - Flow Rate')

flow_76 <- dat %>%
  filter(is.na(kg_per_month)) %>%
  group_by(key) %>%
  filter(!any(param == '00056 - Flow Rate')) %>%
  filter(param == '74076 - Flow')

flow_58 <- dat %>%
  filter(is.na(kg_per_month)) %>%
  group_by(key) %>%
  filter(!any(param == '00056 - Flow Rate')) %>%
  filter(!any(param == '74076 - Flow')) %>%
  filter(param == '00058 - Flow Rate')
N_params<- dat %>%
  filter(is.na(kg_per_month)) %>%
  filter(!param == '00056 - Flow Rate') %>%
  filter(!param == '74076 - Flow') %>%
  filter(!param == '00058 - Flow Rate') %>%
  distinct() %>%
  filter(!is.na(conc_avg)) 

flow_cleaned<-rbind(flow_56,flow_76,flow_58, N_params) 

#review duplicates
flow_cleaned %>%
  filter(is.na(kg_per_month)) %>%
  group_by(key, param_simple)%>%
  tally() %>%
  arrange(desc(n))

#now there are only 3 duplicated rows; 2 of which are for max values
#and this one CT0001180_1_1989-12-31 has true duplicates; we can just remove it

flow_cleaned <- flow_cleaned %>% 
  #removing four problematic observations
  filter(key!='CT0001180_1_1989-12-31') %>%
  filter(key!='CT0001406_3_1989-12-31') %>%
  filter(key!='CT0001406_4_1989-12-31') %>%
  filter(key!='CT0100081_1_1989-12-31') %>% 
   #removing unwanted columns
  select(
    !c('Monitoring Dt (MVDT)',
        'Discharge Designator (DRID)', 
       'Parameter Cd (PRAM)',
       'Monitoring Location (MLOC)',
       '#MULTIVALUE',   
       'Meas Quant Maxm (MQMX)', 
       'Quant Unit Cd (LQUC)',
       'Quant Unit Cd Desc (LQUCD)',
       'Meas Conc Minm (MCMN)',
       'Meas Conc Maxm (MCMX)...15',
       'Meas Conc Maxm (MCMX)...16',
       'Conc Unit Cd (LCUC)',
       'Conc Unit Cd Desc (LCUCD)' )
  )


#subset of data that DO NOT report loads, then
#pivot wider and then remove columns with only NAs (since some parameter/unit combinations that don't make sense)

flow_N_concentration <- flow_cleaned %>%
  group_by(key) %>%
  filter(is.na(kg_per_month))  %>%
   pivot_wider(names_from = param_simple,
              values_from = c(mg_N_per_L, million_L_per_month)) %>%
  select_if(not_all_na) 

#calculating N loads
N_load_b <- flow_N_concentration %>%
  mutate(kg_N_TN_per_month = mg_N_per_L_TN * million_L_per_month_flow_rate) %>%
  mutate(kg_N_ON_per_month = mg_N_per_L_ON * million_L_per_month_flow_rate) %>%
  mutate(kg_N_NH3_per_month = mg_N_per_L_NH3 * million_L_per_month_flow_rate) %>%
  mutate(kg_N_NO2_per_month = mg_N_per_L_NO2 * million_L_per_month_flow_rate) %>%
  mutate(kg_N_NO3_per_month = mg_N_per_L_NO3 * million_L_per_month_flow_rate) %>%
  select(
    -mg_N_per_L_TN,
    -mg_N_per_L_NH3,
    -mg_N_per_L_NO2,
    -mg_N_per_L_NO3,
    -million_L_per_month_flow_rate,
  )

#combine N columns in N_load_b to standardize and join dataframes
names(N_load_b)


N_load_b$kg_N_inorganic_per_month <- N_load_b$kg_N_NH3_per_month +
  N_load_b$kg_N_NO3_per_month +
  N_load_b$kg_N_NO2_per_month

#adding organic N and inorganic N together as 'composite' N
N_load_b$kg_N_sum_per_month <- N_load_b$kg_N_ON_per_month +
  N_load_b$kg_N_inorganic_per_month

#remove intermediate columns we no longer need
N_load_b <- N_load_b %>%
  select(
    -kg_N_NH3_per_month,-kg_N_NO3_per_month,
    -kg_N_NO2_per_month,
    -kg_N_inorganic_per_month,
    -kg_N_ON_per_month
  )

#combine the different N columns; use total N when reported, or the sum of all other
#forms of N if TN is not reported
N_load_b$kg_N_TN_per_month <-
  ifelse(
    !is.na(N_load_b$kg_N_TN_per_month),
    N_load_b$kg_N_TN_per_month,
    N_load_b$kg_N_sum_per_month
  )

#remove N sum column
N_load_b<- N_load_b %>%
  select(-kg_N_sum_per_month)

names(N_load_b)

#for N_load_a
#combine the different N columns; use total N when reported, or the sum of all other
#forms of N if TN is not reported

N_load_a$kg_N_TN_per_month <-
  ifelse(
    !is.na(N_load_a$kg_N_TN_per_month),
    N_load_a$kg_N_TN_per_month,
    N_load_a$kg_N_NH3_per_month
  )

#remove NH3 column
N_load_a<- N_load_a %>%
  select(-kg_N_NH3_per_month)

names(N_load_a)

#combine dataframes
N_load<-rbind(N_load_a,N_load_b)
```

Loading in NPDES database with NPDES ID to match locations and other data.

```{r match locations}
outfall_locations <-
  read_csv(file = here('data', 'NPDES_PERM_FEATURE_COORDS.csv'))

names(outfall_locations)

names(N_load)

#need to remove leading zeroes from outfall column
N_load$Outfall <- sub('^0+', '', N_load$Outfall)
N_load$Outfall

#create key variables with permit and locations for join
N_load$permit_outfall <- paste0(N_load$permit, '_', N_load$Outfall)
outfall_locations$permit_outfall <-
  paste0(outfall_locations$EXTERNAL_PERMIT_NMBR,
         '_',
         outfall_locations$PERM_FEATURE_NMBR)

#join locations based on permit number and outfall number
N_load_out <- left_join(N_load, outfall_locations, by = 'permit_outfall')

length(unique(as.character(N_load$permit_outfall)))
length(unique(as.character(N_load_out$permit_outfall)))
dim(N_load_out)


```

Now join the HUC8 IDs to the data.

```{r}
HUC8_outfall_locations <-
  read_csv(file = here('data', 'HUC8_NPDES_outfall_intersection.csv'))

#select columns
HUC8_outfall_locations <- HUC8_outfall_locations %>%
  select('EXTERNAL_PERMIT_NMBR',
         'PERM_FEATURE_NMBR',
         'huc8',
         'name')

names(HUC8_outfall_locations)

#need to remove leading zeroes from outfall column
HUC8_outfall_locations$PERM_FEATURE_NMBR <- sub('^0+', '', HUC8_outfall_locations$PERM_FEATURE_NMBR)
HUC8_outfall_locations$PERM_FEATURE_NMBR


#create key variables with permit and locations for join
HUC8_outfall_locations$permit_outfall <-
  paste0(HUC8_outfall_locations$EXTERNAL_PERMIT_NMBR,
         '_',
         HUC8_outfall_locations$PERM_FEATURE_NMBR)

#join locations based on permit number and outfall number
dat_join<- left_join(N_load_out, HUC8_outfall_locations, by = 'permit_outfall')

write.csv(dat_join,
          file = here('data', 'PCS_data_clean.csv'),
          row.names = F)

```
